Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach

### 2021-03-29 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Final proof-read and edit before submission

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Examining spatial equity in accessibility to a public bicycle share
program using a balanced floating catchment area approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, 1280 Main Street West,
Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, 1265 Military Trail, Toronto, ON
M1C1A4}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Public bicycle share programs (PBSPs) are implemented to promote cycling
  as a convenient and sustainable mode of transportation. These systems
  can also play a role in addressing transportation needs and advancing
  transportation equity if they make cycling more accessible to lower
  income or disadvantaged populations. In Ontario, the City of Hamilton
  launched a public bicycle share program in 2015 that currently has over
  900 operational bicycles and 130 docking stations. This system was
  expanded in 2018 by an equity initiative that added twelve docking
  stations with the explicit objective of increasing spatial equity in
  access. Since the cost of reaching a bicycle share station decreases the
  potential of accessing the program, the location of stations matters,
  and previous research found that Hamilton's public bicycle share program
  targeted well disadvantaged areas in the city. In this research, we
  revisit the case of Hamilton and investigate differentials in
  accessibility to bicycle share stations using a balanced floating
  catchment area (BFCA) accessibility approach. We compare accessibility
  with and without the equity stations to assess the effect of the
  initiative. We implement our analysis parting from micro zones to better
  reflect walking to a docking station, and conduct a sensitivity analysis
  at several walking time thresholds. We then reaggregate the estimated
  accessibility for further analysis using census data. Our findings
  indicate that the addition of equity stations increased the horizontal
  equity of the system by growing the serviced population at every
  threshold examined. With respect to vertical equity, although
  accessibility increased for the whole population, this increase was
  relatively modest especially for population in the bottom 20\% of median
  total household income.
  \end{abstract}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The potential of public bicycle share programs (PBSPs) to increase
cycling levels is but one of many reasons for implementing such programs
in urban areas (Hosford et al., 2018, 2019). As a healthy, inexpensive,
and convenient form of public transportation, shared bicycles can
encourage individuals to take up cycling for short local trips or first
and last mile trips to other public transportation instead of using
personal vehicles. These systems can also play a role in addressing
transportation needs and advancing transportation equity if they make
cycling more accessible to lower income or disadvantaged populations. As
these systems become increasingly common, their introduction has been
accompanied by a flurry of research that investigates them from the
perspective of spatial and transportation equity (e.g., Smith et al.,
2015; Hosford and Winters, 2018; Hull Grasso et al., 2020; Mooney et
al., 2019; Qian and Jaller, 2020; Qian et al., 2020). Indeed, although
these programs are available to the general public and ought to be
accessible to any individuals who wish to use them, research on PBSPs
located in North American and European cities indicates that inequity in
spatial accessibility needs redressing.

In this paper we investigate the case of the PBSP in the City of
Hamilton, in Ontario, Canada. Hamilton Bike Share (previously also known
as Social Bicycles, or SoBi) was launched in 2015 and currently has over
900 operational bicycles and 130 docking stations. An interesting
feature of this system is that it was expanded by an equity initiative
that introduced twelve docking stations in 2018 with the explicit
objective of increasing spatial accessibility. Hamilton Bike Share was
already studied by Hosford and Winters (2018) as part of a selection of
PBSPs in Canada. These authors found that most of the cities in their
research could benefit of greater efforts to expand service to areas
with lower socio-economic populations - but Hamilton fared somewhat
better than other cities in this respect. What distinguishes our
research is the use of a balanced floating catchment area (BFCA)
accessibility approach that accounts for the supply of stations and the
potential demand from the population serviced. This requires a more
disaggregated approach than the use of census geographies (q.v., Hosford
and Winters, 2018), since many trips to a bike share station likely
happen at a level lower than even the smallest census geography. For
this reason, we implement our analysis parting from micro population
zones to better reflect the friction of walking to a docking station.
Further, we conduct a sensitivity analysis at several walking time
thresholds. In terms of spatial equity, we compare accessibility with
and without the equity stations to assess the effect of the initiative,
before reaggregating the data for further analysis using median total
household income information from the census. We find that the addition
of equity stations does increase the serviced population at every
threshold examined. However, although accessibility increased for the
whole population, this increase was relatively modest, especially for
population in the bottom 20\% of median total household income. The
analysis also identifies areas under-serviced within the core service
area of the system.

This paper is an example of open and reproducible research that uses
only open software for transportation and statistical analysis (Bivand,
2020; Lovelace, 2021). All data were obtained from publicly available
sources and organized in the form a data package. Following best
practices in spatial data science (Brunsdon and Comber, 2020), the code
and data needed to reproduce (or modify/extend) the analysis are
available for download.\footnote{https://drive.google.com/drive/folders/1ZGRpSN2LxS2Fq2tLVJgfm1sI6I26vzZh?usp=sharing}

\hypertarget{literature-review}{%
\section{Literature Review}\label{literature-review}}

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users
(Fishman, 2016). In many cities, males use bike share more than females
({\textbf{???}}; Brey et al., 2017; Nickkar et al., 2019; Reilly, S. M.
Wang, et al., 2020; Winters et al., 2019) as do younger age cohorts
(Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011). However,
one study found that bike share users in Washington, DC were more likely
to be female (Buck et al., 2013), which suggests that the gender gap
among cyclists who use bike share is less disparate than the gap for
personal bicycle use (Fishman, 2016). There is some evidence that bike
share users are less likely to own a car (Buck et al., 2013; Reilly,
Noyes, et al., 2020). However, the relationship between income or
education and bike share use is less clear-cut. Stations in
disadvantaged communities in Chicago have been found to generate most of
the average annual trips (Qian and Jaller, 2020) and individuals from
minority or lower socioeconomic status neighbourhoods in
Minneapolis-St.~Paul used the city's PBSP more (J. Wang and Lindsey,
2019a). Similar findings were reported in London ({\textbf{???}}). Being
university educated was a significant correlate of bike share use in
Montreal, Canada (Fuller et al., 2011). Perhaps not coincidentally,
financial savings have been found to motivate those on a low income to
use bike share (Fishman, 2016).

Many studies have found that geographic proximity to a bicycle share
station is an important determinant of membership and use (Fishman,
2016; Fuller et al., 2011). This makes sense given that individuals are
more likely to use services or programs that they can easily reach.
Several studies have recently explored the equity of PBSPs by primarily
examining who has access to bike share (e.g., differences by
demographics or socioeconomic status) and where stations are located. It
is important to note that equity can be achieved in two different ways:
equal spatial distribution across a region (e.g., horizontal equity) or
greater access for vulnerable or disadvantaged populations (e.g.,
vertical equity) (Chen et al., 2019). Both are of interest to
researchers and transport planners since they are often linked in that
advantage, or conversely disadvantage, has spatial patterns. Using a
negative binomial regression model, Qian and Jaller (2020) estimated
ridership in Chicago's PBSP among disadvantaged communities and found
some disparities. While annual members in disadvantaged communities have
a significantly lower share of trips compared to other areas in the
city, they make longer trips. This suggests that they may be using PBSPs
for utilitarian trips (e.g., commuting), which points to the importance
of ensuring equitable access (Qian and Jaller, 2020). Similar results
were found in Philadelphia, where lower income areas generated fewer
trips and it was suggested that efforts to increase equity within the
program have not been as successful as intended (Caspi and Noland,
2019). In the case of Seattle, all neighbourhoods had some level of
access to dockless bikes but those with higher incomes and more
residents of higher education had more bikes (Mooney et al., 2019).
Babagoli et al.~(2019) also found that neighbourhoods in New York City
with higher affluence had the greatest proportion of Citi Bike stations.
Researchers have recommended targeted expansion of stations to areas
that use bike share but have inequitable access ({\textbf{???}}).

On the whole, existing studies highlight the need for bicycle share
systems to be more highly accessible for diverse populations in order to
increase use beyond the ``typical'' users. This has been the focus of
recent research (see, among others, Auchincloss et al., 2020; Hull
Grasso et al., 2020; MacArthur et al., 2020). Offering more people the
option of using sustainable and active transportation, particularly
those who have lower socioeconomic status and might benefit the most, is
a worthy policy goal for cities with PBSPs. Several cities have launched
specific programs to address barriers to bike share or to expand service
to more deprived areas: London, United Kingdom ({\textbf{???}});
Philadelphia, Pennsylvania (Caspi and Noland, 2019), and Hamilton,
Ontario (Hosford and Winters, 2018). Findings to date suggest that
horizontal equity can be easier to achieve than vertical equity.
However, exploring transportation equity by investigating where bicycle
share stations are located, often using neighbourhoods or census tracts
as the geographical unit of analysis, can ignore or miss the benefits
that may be derived from adjacent zones. Meaning that, stations may be
lacking in certain neighbourhoods but there may be stations accessible
within a reasonable walking time. This is where geographical
accessibility becomes an important consideration.

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved over the past century and has increasingly become
linked with recent interest in prioritizing local proximity and modes
that are suitable for local trips like walking and cycling (Levine,
2020). As a measure of the ease of reaching potential destinations
spread spatially in a given area, accessibility is relevant to PBSPs
because it can identify current inequaltiy in the provision of
infrastructure, as well as guide interventions that increase access for
groups that are under-serviced or address gaps in transportation
options. It also addresses some of the challenges of other performance
measures such as level of service within a transportation network by
measuring person-based indicators and exploring differences in use
between population groups (Páez et al., 2012).

Beyond the utility derived from using shared bikes to destinations of
value, an important aspect of accessibility associated with PBSPs is the
distance an individual must travel to reach a bicycle share station
(Kabra et al., 2020; J. Wang and Lindsey, 2019b). Since the time or
distance needed to reach a bicycle share station decreases the potential
of accessing the program and ultimately their use of the program, the
location and size (e.g., number of bicycle racks) of stations matters.
Indeed, distance to bicycle share stations is associated with use
(Fuller et al., 2011; J. Wang and Lindsey, 2019b) and can be a barrier
to using PBSPs (Fishman et al., 2014). Kabra et al.~(2020) found that
the majority of bike share usage in Paris comes from areas within 300 m
of stations, which amounts to 2-4 minutes walking by an able-bodied
adult. Furthermore, similar to other public amenities affected by
crowding, like health care services (e.g., Pereira, Carlos Kauê Vieira
Braga, et al., 2021), the utility of stations is limited by the number
of bicycles that they can hold. The program may not necessarily be
improved if stations are easy to reach but offer only a small number of
bicycles. Likewise, more people may not opt to use the program if the
supply of bicycles available at nearby stations is insufficient to meet
demand. The location and size of stations is important to increase the
utility of this public transportation option for more people, thus
achieving vertical equity. Accessibility analyses for PBSPs constitute a
positive and evaluation-based approach that also has the potential to
inform equity efforts. For instance, Wang and Lindsey (2019b)
investigated whether new or relocated bicycle share stations increased
accessibility and use, which offered important insights to improve the
performance of the program.

@@

In contrast, floating catchment area (FCA) methods incorporate
information on capacity and demand in calculating accessibility. FCA
measures have been widely employed in healthcare accessibility research
and are typically calculated across two steps. In the first, a ratio of
supply to demand at service locations is calculated, such as the number
of beds at a hospital divided by the number of people within the
catchment area of the hospital, weighted by the distance involved in
reaching the facility. Next, these service level ratios are allocated
back to the population centres and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. While there have been many
methodological innovations in FCA methods (for example, {\textbf{???}};
{\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}), a
recent improvement to this approach was achieved through a simple and
intuitive balancing of the impedance that addressed the effects of
demand and service inflation found in earlier FCA approaches (see Paez
et al., 2019).

When measuring accessibility, researchers have taken different
approaches when it comes to aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregate data from entire
neighbourhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals, which is useful for addressing vertical
inequity in PBSP usage.

Using the balanced floating catchment area method (Paez et al., 2019), a
novel approach that has not been used yet in cycling research, with
disaggregate population-level data, in this paper we examine
accessibility to a public bicycle share program in Hamilton, Ontario. We
(1) conduct a sensitivity analysis by measuring accessibility and level
of service to bike share stations at different walking time thresholds
to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
(2) explore the contribution of specific stations that were added to
Hamilton's PBSP to reducing both horizontal and vertical inequities; and
(3) examine whether disparities in accessibility exist according to
median total household income of dissemination areas within the core
service area.

\hypertarget{sec:study}{%
\section{Case Study}\label{sec:study}}

\hypertarget{original-system}{%
\subsection{Original System}\label{original-system}}

The focus of this study is the city of Hamilton, located in Ontario,
Canada. The city launched a public bicycle share program in March 2015
with 115 stations and 750 bicycles (Hamilton, 2015). Before June 2020,
the program was known as Social Bicycles or SoBi Hamilton but is now
called Hamilton Bike Share. Stations are spaced between 300 and 600
metres apart (Scott et al., 2021). The core service area spans 40 sq.km
of the city although it was planned to be 20 sq.km ({\textbf{???}}), and
roughly 138,000 people are within 30 minutes walking of a bike share
station {[}see Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This
represents roughly one fifth of the total population of the Hamilton
Census Metropolitan Area according to the 2016 Canadian Census. The City
of Hamilton undertook a large public engagement campaign to validate the
locations of bicycle share stations that had been selected and to
crowdsource potential locations for additional locations
({\textbf{???}}). Most of the suggested locations were in the east end
of the core service area that lack stations or in neighbourhoods not
serviced by the PBSP. The program was enthusiastically welcomed in the
city in 2015 - within three weeks of launching, 10,000 trips had been
made (Hamilton, 2015), however inadequate coverage given the size of the
Hamilton Bike Share's service area was identified as a problem early on,
and transportation planners noted that the small size (i.e., supply of
bicycle racks) and low quantity of stations would lead to challenges in
balancing the system ({\textbf{???}}). Based on public feedback, 13
stations were added to the original system.

\hypertarget{equity-initiative}{%
\subsection{Equity Initiative}\label{equity-initiative}}

In 2017, Hamilton Bike Share Inc., the non-profit organization that
operates the program, initiated an equity program, Everyone Rides
Initiative (ERI), with the objective of reducing barriers that may
prevent individuals from accessing bike share in Hamilton. Additional
bicycles and stations were added to the program which expanded it to
more disadvantaged areas in the core service area {[}see Figure 2{]}.
The program also offers subsidized memberships to individuals who
identify as low income, and complements these initiatives with cycle
skills education. A comparable program can be found in Philadelphia (see
Caspi and Noland, 2019).

\hypertarget{current-system}{%
\subsection{Current System}\label{current-system}}

As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations {[}see
Figure \ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
memberships (Hamilton, 2015). The core service area remains 40 sq.km.
The PBSP has 13 stations that were added as part of the ERI initiative;
we refer to these as ``equity stations'' throughout the paper while all
other stations are ``conventional stations''. In total, 117 stations are
``conventional'' and 13 are ``equity''.

\hypertarget{membership}{%
\subsection{Membership}\label{membership}}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Examining horizontal and vertical equity to a public bicycle share
program using a balanced floating catchment area accessibility approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
Street West, Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Public bicycle share programs (PBSPs) can play a role in advancing
  transportation equity if they make cycling more accessible to
  disadvantaged populations. In Ontario, the City of Hamilton expanded
  their PBSP in 2018 by adding twelve docking stations with the explicit
  objective of increasing spatial equity in access. In this case study, we
  investigate differentials in accessibility to stations using a balanced
  floating catchment area accessibility approach and compare accessibility
  with and without the equity stations. We analyze micro zones to better
  reflect walking to a station and conduct a sensitivity analysis at
  several walking time thresholds. We then reaggregate the estimated
  accessibility for further analysis using census data. Our findings
  indicate that equity stations increased the serviced population at every
  threshold examined. Although accessibility increased for the whole
  population, this increase was relatively modest especially for
  population in the bottom 20\% of median total household income.
  \end{abstract}

@@

\hypertarget{introduction}{%
\section{1. Introduction}\label{introduction}}

The potential of public bicycle share programs (PBSPs) to increase
cycling levels is but one of many reasons for implementing such programs
in urban areas (Hosford et al., 2018, 2019). As a healthy, inexpensive,
and convenient form of public transportation, shared bicycles can
encourage individuals to take up cycling for short local trips or first
and last mile trips to other public transportation instead of using
personal vehicles. These programs can also play a role in addressing
transportation needs and advancing transportation equity if they make
cycling more accessible to individuals and groups with lower
socioeconomic status. As these programs become increasingly common,
their introduction has been accompanied by a flurry of research that
investigates them from the perspective of spatial and transportation
equity (e.g., Hosford and Winters, 2018; Hull Grasso et al., 2020;
Mooney et al., 2019; Qian and Jaller, 2021, 2020; Smith et al., 2015).
Indeed, although these programs are available to the general public and
ought to be accessible to any individual who wishes to use them,
research on PBSPs located in North American and European cities
indicates that inequity in spatial accessibility needs redressing.

In this paper, we investigate the case of the PBSP in the city of
Hamilton, in Ontario, Canada. Hamilton Bike Share was launched in 2015
and currently has over 900 operational bicycles and 130 docking
stations. An interesting feature of this program is that it was expanded
by an equity initiative that introduced twelve stations in 2018 with the
explicit objective of increasing spatial accessibility. Hamilton Bike
Share was already studied by Hosford and Winters (2018) as part of a
selection of PBSPs in Canada. The authors found that most of the cities
in their research could benefit of greater efforts to expand service to
areas with lower socioeconomic groups - but Hamilton fared somewhat
better than other cities in this respect. What distinguishes our
research from existing literature is the use of a balanced floating
catchment area (BFCA) accessibility approach that accounts for the
supply of stations and the potential demand from the population
serviced. This requires a more disaggregated approach than the use of
census geographies (q.v., Hosford and Winters, 2018), since walking
trips to a bicycle share station likely happen at a level lower than
even the smallest census geography. For this reason, we implement our
analysis parting from micro population zones to better reflect the
friction of walking to a docking station, which is an important
component of a bike share trip (Chen et al., 2019). Further, we conduct
a sensitivity analysis at several walking time thresholds. In terms of
spatial equity, we compare accessibility with and without the equity
stations to assess the effect of the initiative, before reaggregating
the data for further analysis using median total household income
information from the census.

This paper is an example of open and reproducible research that uses
only open software for transportation and statistical analysis (Bivand,
2020; Lovelace, 2021). All data were obtained from publicly available
sources and organized in the form of a data package. Following best
practices in spatial data science (Brunsdon and Comber, 2020), the code
and data needed to reproduce, modify or extend the analysis are
available for download.\footnote{https://github.com/paezha/Accessibility-Sobi-Hamilton}

\hypertarget{literature-review}{%
\section{2. Literature Review}\label{literature-review}}

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users
(Fishman, 2016). In many cities, males use bike share more than females
(Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
Reilly, S. M. Wang, et al., 2020; Winters et al., 2019) as do younger
age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011).
However, one study found that bike share users in Washington, DC were
more likely to be female (Buck et al., 2013), which suggests that the
gender gap among cyclists who use bike share is less disparate than the
gap for personal bicycle use (Fishman, 2016). There is some evidence
that bike share users are less likely to own a car (Buck et al., 2013;
Reilly, Noyes, et al., 2020). However, the relationship between income
or education and bike share use is less clear-cut. Stations in
disadvantaged communities in Chicago have been found to generate most of
the average annual trips (Qian and Jaller, 2020) and individuals from
minority or lower socioeconomic status neighborhoods in
Minneapolis-St.~Paul used the city's PBSP more (J. Wang and Lindsey,
2019a). Similar findings were reported in London (Ogilvie and Goodman,
2012). Being university educated was a significant correlate of bike
share use in Montreal, Canada (Fuller et al., 2011). Perhaps not
coincidentally, financial savings have been found to motivate those on a
low income to use bike share (Fishman, 2016).

Many studies have found that geographic proximity to a bicycle share
station is an important determinant of membership and use (Fishman,
2016; Fuller et al., 2011). This makes sense given that individuals are
more likely to use services or programs that they can easily reach.
Several studies have recently explored the equity of PBSPs in North
America by primarily examining who has access to bike share (e.g.,
differences by demographics or socioeconomic status) and where stations
are located. It is important to note that equity can be achieved in two
different ways: balanced or equal spatial distribution across a region
for all similar groups (e.g., horizontal equity) or greater and targeted
access to accommodate vulnerable or disadvantaged populations (e.g.,
vertical equity) (see Chen et al., 2019). Both are of interest to
researchers and transportation planners since they are often linked in
that advantage, or conversely disadvantage, has spatial patterns. Using
a negative binomial regression model, Qian and Jaller (2020) estimated
ridership in Chicago's PBSP and found some disparities. A minority of
bike share stations were found to be located in disadvantaged
communities, while annual members from disadvantaged communities have a
lower share of trips compared to other areas in the city. This suggests
that they may be using PBSPs for utilitarian trips (e.g., commuting),
which points to the importance of ensuring equitable access (Qian and
Jaller, 2020). Similar results were found in Philadelphia. Despite
efforts to increase equity within the city's PBSP, census block groups
with lower median income generated fewer trips which suggested that
efforts have not been as successful as intended (Caspi and Noland,
2019). Trips from stations in such areas are utilitarian (e.g.,
commuting to work), which points to the importance of ensuring equitable
access (Caspi and Noland, 2019). In the case of Seattle, all
neighborhoods had some level of access to dockless bicycles but those
with higher incomes and more residents of higher education had more
bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
neighborhoods in New York City with higher affluence had the greatest
proportion of Citi Bike stations.

On the whole, existing studies highlight the need for PBSPs to be more
highly accessible for diverse populations in order to increase use
beyond the ``typical'' users. This has been the focus of recent research
(see, among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. A few North American cities have launched
specific programs to address barriers to bike share or to expand service
to more deprived areas: Philadelphia, Pennsylvania (Caspi and Noland,
2019), and Hamilton, Ontario (Hosford and Winters, 2018). However,
exploring transportation equity by investigating where bicycle share
stations are located, often using neighborhoods or census tracts as the
geographical unit of analysis, can ignore or miss the benefits that may
be derived from adjacent zones. Meaning that, stations may be lacking in
certain neighborhoods but there may be stations accessible within a
reasonable walking time. This is where geographical accessibility
becomes an important consideration.

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved over the past century and has increasingly become
linked with recent planning interests in prioritizing modes that are
suitable for local trips like walking and cycling (Levine, 2020). As a
measure of the ease of reaching potential destinations spread spatially
in a given area, accessibility is relevant to PBSPs because it can
identify current inequality in the provision of infrastructure, as well
as guide interventions that increase access for groups that are
under-serviced or address gaps in transportation options. It also
addresses some of the challenges of other performance measures such as
level of service within a transportation network by measuring
person-based indicators and exploring differences in use between
population groups (Páez et al., 2012).

Beyond the utility derived from using shared bikes to destinations of
value, an important aspect of accessibility associated with PBSPs is the
distance an individual must travel to reach a bicycle share station
(Kabra et al., 2020; J. Wang and Lindsey, 2019b). Since the time or
distance needed to reach a bicycle share station decreases the potential
of accessing the program and ultimately their use of the program, the
location and size (e.g., number of bicycle racks) of stations matters.
Indeed, distance to bicycle share stations is associated with use
(Fuller et al., 2011; J. Wang and Lindsey, 2019b) and can be a barrier
to using PBSPs (Fishman et al., 2014). Kabra et al.~(2020) found that
the majority of bike share usage in Paris comes from areas within 300 m
of stations, which amounts to 2-4 minutes walking by an able-bodied
adult. Furthermore, similar to other public amenities affected by
crowding, like health care services (e.g., Pereira, Carlos Kauê Vieira
Braga, et al., 2021), the utility of stations is limited by the
\emph{maximum} number of bicycles that they can hold. The program may
not necessarily be improved if stations are easy to reach but offer only
a small number of bicycles. Likewise, more people may not opt to use the
program if the supply of bicycles available at nearby stations is
insufficient to meet demand. The location and size of stations is
important to increase the utility of this public transportation option
for more people, thus achieving vertical equity. Accessibility analyses
for PBSPs constitute a positive and evaluation-based approach that also
has the potential to inform equity efforts. For instance, Wang and
Lindsey (2019b) investigated whether new or relocated bicycle share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

@@

In contrast, floating catchment area (FCA) methods incorporate
information on capacity and demand in calculating accessibility. FCA
measures have been widely employed in healthcare accessibility research
and are typically calculated across two steps. In the first, a ratio of
supply to demand at service locations is calculated, such as the number
of beds at a hospital divided by the number of people within the
catchment area of the hospital, weighted by the distance involved in
reaching the facility. Next, these service level ratios are allocated
back to the population centers and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. While there have been many
methodological innovations in FCA methods (for example, Delamater, 2013;
Luo and Qi, 2009; Luo and Wang, 2003; Radke and Mu, 2000; Wan et al.,
2012), a recent improvement to this approach was achieved through a
simple and intuitive balancing of the impedance that addressed the
effects of demand and service inflation found in earlier FCA approaches
(see Paez et al., 2019).

When measuring accessibility, researchers have taken different
approaches when it comes to aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals, which is useful for addressing vertical
inequity in PBSP usage. Chen et al.~(2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation''.

Using the balanced floating catchment area method (Paez et al., 2019), a
novel approach that has not been used yet in cycling research, with
disaggregated population data, we examine accessibility to a public
bicycle share program in Hamilton, Ontario. In this paper, we (1)
conduct a sensitivity analysis by measuring accessibility and level of
service to bicycle share stations at different walking time thresholds
to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
(2) explore the contribution of specific stations that were added to
Hamilton's PBSP to reducing both horizontal and vertical inequities; and
(3) examine whether disparities in accessibility exist according to
median total household income of dissemination areas within the core
service area.

\hypertarget{sec:study}{%
\section{3. Case Study}\label{sec:study}}

\hypertarget{original-system}{%
\subsection{3.1. Original System}\label{original-system}}

The focus of this study is the city of Hamilton, located in Ontario,
Canada. The city launched a public bicycle share program in March 2015
with 115 stations and 750 bicycles (Hamilton, 2015a). Before June 2020,
the program was known as Social Bicycles or SoBi Hamilton, but is now
called Hamilton Bike Share. Stations are spaced between 300 and 600 m
apart (Scott et al., 2021). The core service area spans 40 sq.km of the
city although it was planned to be 20 sq.km (Hamilton, 2015b), and
roughly 138,000 people are within 30 minutes walking of a bike share
station {[}see Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This
represents roughly one fifth of the total population of the Hamilton
Census Metropolitan Area according to the 2016 Canadian Census. The City
of Hamilton undertook a large public engagement campaign to validate the
locations of bicycle share stations that had been selected and to
crowdsource potential locations for additional locations (Hamilton,
2014). Most of the suggested locations were in the east end of the core
service area that lack stations or in neighborhoods not serviced by the
PBSP. The program was enthusiastically welcomed in the city in 2015 -
within three weeks of launching, 10,000 trips had been made (Hamilton,
2015a), however inadequate coverage given the size of Hamilton Bike
Share's service area was identified as a problem early on, and
transportation planners noted that the small size (i.e., supply of
bicycle racks) and low quantity of stations would lead to challenges in
balancing the system (Hamilton, 2015b). Based on public feedback,
thirteen stations were added to the original system.

\hypertarget{equity-initiative}{%
\subsection{3.2. Equity Initiative}\label{equity-initiative}}

In 2017, Hamilton Bike Share Inc., the non-profit organization that
operates the program, initiated an equity program, Everyone Rides
Initiative (ERI), with the objective of reducing barriers that may
prevent individuals from accessing bike share in Hamilton. Additional
bicycles and stations were added to the program which expanded it to
more disadvantaged areas in the core service area. The program also
offers subsidized memberships to individuals who identify as low income,
and complements this service with cycle skills education. A comparable
program can be found in Philadelphia (see Caspi and Noland, 2019).

\hypertarget{current-system}{%
\subsection{3.3. Current System}\label{current-system}}

As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations {[}see
Figure \ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
memberships (Hamilton, 2015a). The core service area remains 40 sq.km.
The PBSP has twelve stations that were added as part of the ERI
initiative; we refer to these as ``equity stations'' throughout the
paper while all other stations are ``conventional stations''. In total,
118 stations are ``conventional'' and 12 are ``equity''.

\hypertarget{membership}{%
\subsection{3.4. Membership}\label{membership}}
